---
title: "POLS6382 Methods III HM1 Review"
author: "Ling Zhu and Emily Lee"
date: "2025-09-24"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Overall Comments on HM1

-   Clean/well organized data reports! Everyone did a good job!

-   Replicability of R code: Some students reported R code that are not consistent with the reported statistical results. 

-   Markdown presentation: R Code-Chunk setting, "warning=FALSE, message=FALSE, echo=TRUE"

## Question 1

-   The issue of losing country names, when loading data using package \`\`haven."
-   Scatter plot: Full sample vs. by country.



## Question 2 Grid Search

\footnotesize

-   Poisson Likelihood vs. Poisson log-Likelihood

-   Data: $y_i=(2,2,4,1,0,1,3,2)$

-   $\sum y_i =15$

-   $\prod y_i!=1152$

-   The likelihood is: $L(\lambda | y_1, y_2...y_8) =\frac{e^{-8\lambda} \lambda^{\Sigma y_i}}{\prod y_i!}$

-   The Log-Likelihood is: $Log[\frac{e^{ -8\lambda} \lambda^{\Sigma y_i}}{\prod y_i!}]=-8\lambda+log(\lambda^{15})-log(\prod y_i!)$

```{r echo=TRUE, warning=FALSE}
data<-c(2,2,4,1,0,1,3,2)
lambda <- seq(0, 4, 0.01)
# Likelihood Function
likelambda <- (exp(-8 * lambda) * lambda^15)/1152
# Log-Likelihood Function
loglike<--8 * lambda + log(lambda^15) - log(1152)
```

-   They actually produce the same solution of $\lambda$ in this case!

-   When plotting the grid-search results, make sure that you label the Y-axis correctly.
